\pkg{dataMaid}: your assistant for documenting supervised data quality screening in \proglang{R}

### 2017-11-09 latex/article_vol3.Rnw
Minor changes regarding typos etc and how figures are included.

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Anne Helby Petersen, Claus Thorn Ekstr\o m} %% comma-separated
\Plaintitle{{dataMaid}: your assistant for data documentation in R} %% without formatting
\Shorttitle{\pkg{dataMaid}: your assistant for data documentation in R} %% a short title (if necessary)

@@

A number of \proglang{R} packages made for other pre-analysis steps
are already available, including \pkg{janitor} \citep{janitor},
\pkg{assertive} \citep{assertive}, \pkg{dplyr} \citep{dplyr},
\pkg{tidyr}\citep{tidyr}, \pkg{data.table} \citep{data.table},
\pkg{DataCombine} \citep{DataCombine}, \pkg{validate}
\citep{validate}, and \pkg{assertr} \citep{assertr}. These packages
focus on different stages of the pre-analysis work.  \pkg{janitor}
provides tools for data import with a particular emphasis on the
challenges of getting neat data frames from Microsoft Excel data
files. \pkg{dplyr}, \pkg{tidyr}, \pkg{data.table} and
\pkg{DataCombine} go a few steps further by providing a wide array of
extremely powerful tools for data wrangling, including a number of
particularly useful functions for merging and working with very large
datasets. When it comes to actual data cleaning, however, the options
are fewer.  \pkg{validate} (and the similar packages \pkg{editrules}
\citep{editrules} and \pkg{deducorrect} \citep{deducorrect} from the
same authors) and \pkg{assertive} offers tools for identifying errors
in a dataset by checking the state of the variable given a set of
pre-specified rules, and their focus is on internal validity rather
than general data screening. In practice, this means that quite elegant
tools for, e.g., linear restraints among the variables in a dataset
can be applied, but looking for potentially miscoded missing values is
not really feasible. The main difference between these two challenges
is the direction in which the data is inspected: While linear
constraints work observation-wise with no ambiguity, determining
whether or not something is a miscoded missing value often requires
knowledge about the full variable (e.g. range or data type), and thus
it should be performed variable-wise. \pkg{validate} does not
currently allow for user-defined extensions of the latter type,
thereby limiting its data cleaning potential.  Automatic data
correction functions are also provided by \pkg{validate} which we
consider to be quite a dangerous cocktail: all power is given to the
the computer with no human supervision, and investigators are less
likely to make an active, case-specific choice regarding the handling
of the potential errors.  Finally, no tools have been made to easily
document exactly which checks and preliminary results were used in the
data cleaning process. The \pkg{assertr} package provide very similar
--- and very nice --- tools to those of \pkg{validate}, but without
any amibitions of conducting auto-cleaning.

@@

One last package that should be mentioned in this context is
\pkg{DataExplorer} \citep{DataExplorer}. While this package does not
address data cleaning issues \emph{per se}, its general strategy is
quite similar to that of \pkg{dataMaid} and to the paradigms presented
below. This package provides a few simple, but practical, tools for
exploratory data analysis, including automated
documentation. Therefore, we find \pkg{DataExplorer} to be a good
candidate for a next-step package after data cleaning is finished.

This manuscript is structured as follows: First, in Section
\ref{sec:usingdataMaid}, we introduce the main representative of the first
paradigm, namely the \code{makeDataReport()} function, which generates data
overview documents. In the \pkg{dataMaid} package, we have
provided a number of default generic checks that cover the data
cleaning challenges we find to be most common and these are also summarized in Section \ref{sec:usingdataMaid}. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{dataMaid}, as motivated
by the second paradigm above. Next, we show step-by-step how the data report and the interactive mode of \pkg{dataMaid} can be combined to conduct a well-documented, systematic data cleaning in Section \ref{sec:bigExample}. Here, we assess and clean a dirty dataset with information about the US presidents.  At last, in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning- and documentation challenges and how \pkg{dataMaid} can be used to
solve them.

%Please note that this description of what \pkg{dataMaid} contains is
%usefully supplemented by a tutorial into how it can be extended. Every
%dataset is different, and some datasets might include problems that
%cannot be detected by our data checking functions. 
%Therefore,
\pkg{dataMaid} was designed to be easily extendend with user-supplied
functions for summarizing, viusalizing and checking data. We have
provided a vignette in which we describe how \pkg{dataMaid} extensions
can be made, such that they are integrate with the
\code{makeDataReport()} function and with the other tools available in
\pkg{dataMaid}.

@@

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=4]{dataMaid_toyData.pdf}}
\frame{\includegraphics[width=7.5cm,page=5]{dataMaid_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\caption{The third and fourth pages of the report created by running \code{makeDataReport()} on the \code{toyData}
  dataset. Here, we see a description of each variable in the dataset, consiting of a summary table, a visualization and a indication of what problems were flagged for the variable (if any). At last, a few lines of metadata about the \code{makeDataReport()} are included for enhancing reproducibility. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\label{fig:example2}
\end{figure}

@@

By default, an \proglang{R} markdown file and a rendered pdf or word
overview document is produced, saved to the working directory and
opened for immediate inspection. Such a data report consists of three
parts, two of which are presented in Figure~\ref{fig:example1}. First,
an overview of what was done is presented under the title \textit{Data
  report overview}. Secondly, an index listing each variable along
with an indication of whether it was found to be problematic or not is
provided. Thirdly, as seen in Figure~\ref{fig:example2}, each variable
in the dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution when relevant and a list of flagged
issues, if any. For instance, as shown in Figure~\ref{fig:example2},
for the \code{numeric}-type variable \code{events} from
\code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and
\code{NaN}), while two values were also flagged as potential outliers
that should be investigated more carefully.

@@

\smallskip Control output and post-processing \\
\quad \code{output} & Type of output file to be produced (html, word (.docx) or pdf) & \code{NULL} (use pdf if \LaTeX\ is found, otherwise word (if on windows), or html)\\
\quad \code{render} & Should the output file be rendered from markdown? & \code{TRUE} \\
\quad \code{openResult} & If a  pdf/html file is rendered, should it
                       automatically open afterwards, and if not,
                       should the \code{rmarkdown} file be opened? & \code{TRUE} \\
\quad \code{replace} & Overwrite an existing file with the same name? & \code{FALSE} \\
\quad \code{vol} & Add a suffix to the file name of the outputted report &   \code{""} (no suffix)\\

@@

In summary, and as indicated in Figure~\ref{figure:cleanStructure}, there are two stages
where \code{makeDataReport()} applies functions to each of the variables:
\begin{enumerate}
\item In the pre-check stage.
\item As part of the summarize/visualize/check (SVC) steps.
\end{enumerate}
Each of these stages are controllable using appropriate function
arguments in \code{makeDataReport()}, and above we have shown examples
of how to tweak them to modify the data cleaning outputs. However, if
the dataset at hand requires new, additional checks, then more control
is needed a package vignette explains the details of how to modify and
expand the possibilities by producing new summary, visual, and check
functions.

@@

Figure \ref{fig:bigExampleP45} presents the remaining two pages with variable presentations. On pages 4 and 5, we find a few remarks:
\begin{itemize}
\item In the birthday variable, there is an entry with the date March 1, 1300 which is a bit of an outlier. 
\item Among the states in which the presidents were born, New York was mistakenly spelled with a lower case "Y" in at least one entry.
\item The variable concerning assassination attempts is coded as a numeric, but the default \code{smartNum = TRUE} setting of \code{makeDataReport()} implies that such a numeric variable with only a few (less that 5) unique values is treated as a factor variable in the data report, thereby providing more relevant summaries, visualizations and checks. This is remarked in the variable presentation for \code{assassinationAttempt} and it can also be seen by the visualization being a barplot rather than a histogram.
\item The variable concerning the sex of the president was skipped, as there is nothing to present when all US presidents have been male so far.
\item The report flags the variable \code{ethnicity} to have suspiciously few observations in one category, "African American". 
\item A few presidents were found to have odd values in the variable describing the duration of their presidencies: Some had very short (outlier) presidencies of less than two years and one was registered to have an \textit{infinite} presidency. 
\item The variable concerning age at inauguration was coded as a character variable, but consists exclusively of numbers and takes a lot of different values and therefore, it was flagged as a potentially misclassified numeric variable.
\item It seems as if one or more presidents have complex numbers as their favorite numbers. As \code{complex} is not a supported variable type in \code{dataMaid} and no strategy for handling this class was provided in the \code{treatXasY} argument, the variable is simply flagged as non-supported.
\end{itemize}
A lot of these mistakes are easily fixable, and we will do so below. However, some of them require more delicate knowledge of the subject matter. For instance, \code{ethnicity} is very reasonably marked as a potentially problematic variable as it includes only a single observation of "African American". However, a human reading this report will know that this does \textit{not} reflect a mistake in the data, but rather a peculiarity in the real world, and as such, it should not be cleaned out.

@@

Next up is the unlikely US president birth day of March 1, 1300. In order to understand if this is a generally problematic observation, or if it is just a mistyped observation, we inspect the full data entry for this person. We can do this using the usual \proglang{R} selection syntax as above, or we can use the value of a \code{identifyOutliers} call to select this observation:

@@

We see that Obama is listed as being president forever, which history has proven to be wrong. Trump, on the other hand, has a missing value for his presidency duration, which is in fact reasonable as we cannot know how long it will be yet (anno 2017). Presidents William Harrison, Zachary Taylor, James Garfield, Warren Harding and Gerald Ford were identified to have very brief presidencies, but these are not mistakes, as any US history textbook can tell us. Thus, the only mistake left to fix is Obama's infinite presidency:

@@

\subsection[Incorporating dataMaid in an automated workflow]{Incorporating \pkg{dataMaid} in automated workflow}
The default settings of \code{makeDataReport()} have been set to facilitate easy and quick data report generation, but unfortunately, this also means that it is not ideal for a more programming-oriented workflow, where the function might not be called by a human. For instance, one might be interested in automatically running \code{makeDataReport()} on all datasets received from a certain client, perhaps via email or through a web upload, and returning a data report for them to inspect and comment before ever looking at the data. In this scenario, there are a few issues with the standard data report:
\begin{enumerate}
\item After rendering, the report is automatically opened. This is not very useful, if the processes are supposed to run in the background.
\item The report generation writes messages to the console while producing and rendering the report.
\item Unless specifically told otherwise, every report created for the same dataset (or different datasets with the same storage name in \proglang{R}) will have the same file name.
\end{enumerate}
Please note that the data report does contain information about who, when and how concerning its generation, so even though the default choices for file names do not make it easy to tell different reports for the same dataset apart, it should be rather easy when inspecting the report manually. These three problems can easily be solved by use of the arguments of \code{makeDataReport()}. Whether or not the outputted file is opened can be controlled through the argument \code{open}. How much information is printed in the console can be adjusted by using the argument \code{quiet}. And convenintely introducing small alterations of the file names can be obtained by use of the \code{vol} argument. For instance, we can make a data report for \code{toyData} that is not opened automatically, produces no output to the console and includes the date and time of its creation in the file name:

@@

\subsection[Using dataMaid for problem flagging]{Using \pkg{dataMaid} for problem flagging}
If the dataset is large and the time available for reading through the data report is scarce, it can be convenient to only make a report concerning the variabels that were flagged to be problematic. This can be achieved by using the \code{onlyProblematic} argument for
\code{makeDataReport()}. By specifying \code{onlyProblematic = TRUE}, only
variables that raise a flag in the checking steps will be summarized
and visualized. But perhaps we are not even interested in obtaining
general information about these variables, but only in getting a quick
overview of the problems they might have. This is obtained by
using the \code{mode} argument:

>>>>

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Anne Helby Petersen, Claus Thorn Ekstr\o m} %% comma-separated
\Plaintitle{{dataMaid}: your assistant for documenting supervised data quality screening in R} %% without formatting
\Shorttitle{\pkg{dataMaid}: your assistant for data documentation in R} %% a short title (if necessary)

@@

A number of \proglang{R} packages made for other pre-analysis steps
are already available, including \pkg{janitor} \citep{janitor},
\pkg{assertive} \citep{assertive}, \pkg{dplyr} \citep{dplyr},
\pkg{tidyr}\citep{tidyr}, \pkg{data.table} \citep{data.table},
\pkg{DataCombine} \citep{DataCombine}, \pkg{validate}
\citep{validate}, and \pkg{assertr} \citep{assertr}. These packages
focus on different stages of the pre-analysis work.  \pkg{janitor}
provides tools for data import with a particular emphasis on the
challenges of getting neat data frames from Microsoft Excel data
files. \pkg{dplyr}, \pkg{tidyr}, \pkg{data.table} and
\pkg{DataCombine} go a few steps further by providing a wide array of
extremely powerful tools for data wrangling, including a number of
particularly useful functions for merging and working with very large
datasets. When it comes to actual data cleaning, however, the options
are fewer.  \pkg{validate} (and the similar packages \pkg{editrules}
\citep{editrules} and \pkg{deducorrect} \citep{deducorrect} from the
same authors) and \pkg{assertive} offer tools for identifying errors
in a dataset by checking the state of the variable given a set of
pre-specified rules, and their focus is on internal validity rather
than general data screening. In practice, this means that quite elegant
tools for, e.g., linear restraints among the variables in a dataset
can be applied, but looking for potentially miscoded missing values is
not really feasible. The main difference between these two challenges
is the direction in which the data is inspected: While linear
constraints work observation-wise with no ambiguity, determining
whether or not something is a miscoded missing value often requires
knowledge about the full variable (e.g. range or data type), and thus
it should be performed variable-wise. \pkg{validate} does not
currently allow for user-defined extensions of the latter type,
thereby limiting its data cleaning potential.  Automatic data
correction functions are also provided by \pkg{validate} which we
consider to be quite a dangerous cocktail: all power is given to the
the computer with no human supervision, and investigators are less
likely to make an active, case-specific choice regarding the handling
of the potential errors.  Finally, no tools have been made to easily
document exactly which checks and preliminary results were used in the
data cleaning process. The \pkg{assertr} package provide very similar
--- and very nice --- tools to those of \pkg{validate}, but without
any amibitions of conducting auto-cleaning.

@@

One last package that should be mentioned in this context is
\pkg{DataExplorer} \citep{DataExplorer}. While this package does not
address data cleaning issues \emph{per se}, its general strategy is
quite similar to that of \pkg{dataMaid} and to the paradigms presented
above. This package provides a few simple, but practical, tools for
exploratory data analysis, including automated
documentation. Therefore, we find \pkg{DataExplorer} to be a good
candidate for a next-step package after data cleaning is finished.

This manuscript is structured as follows: First, in Section
\ref{sec:usingdataMaid}, we introduce the main representative of the first
paradigm, namely the \code{makeDataReport()} function, which generates data
overview documents. In the \pkg{dataMaid} package, we have
provided a number of default generic checks that cover the data
cleaning challenges we find to be most common and these are also summarized in Section \ref{sec:usingdataMaid}. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{dataMaid}, as motivated
by the second paradigm above. Next, we show step-by-step how the data report mode and the interactive mode of \pkg{dataMaid} can be combined to conduct a well-documented, systematic data cleaning in Section \ref{sec:bigExample}. Here, we assess and clean a dirty dataset with information about the US presidents.  At last, in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning- and documentation challenges and how \pkg{dataMaid} can be used to
solve them.

%Please note that this description of what \pkg{dataMaid} contains is
%usefully supplemented by a tutorial into how it can be extended. Every
%dataset is different, and some datasets might include problems that
%cannot be detected by our data checking functions. 
%Therefore,
\pkg{dataMaid} was designed to be easily extendend with user-supplied
functions for summarizing, viusalizing and checking data. In the package, we have
provided a vignette in which we describe how \pkg{dataMaid} extensions
can be made, such that they are integrate with the
\code{makeDataReport()} function and with the other tools available in
\pkg{dataMaid}.

@@

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=4]{dataMaid_toyData.pdf}}
\frame{\includegraphics[width=7.5cm,page=5]{dataMaid_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\caption{The third and fourth pages of the report created by running \code{makeDataReport()} on the \code{toyData}
  dataset. Here, we see a description of each variable in the dataset, consisting of a summary table, a visualization and an indication of what problems were flagged for the variable (if any). At last, a few lines of metadata about the \code{makeDataReport()} are included for enhancing reproducibility. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\label{fig:example2}
\end{figure}

@@

By default, an \proglang{R} markdown file and a rendered pdf,  word or html
overview document is produced, saved to the working directory and
opened for immediate inspection. Such a data report consists of three
parts, two of which are presented in Figure~\ref{fig:example1}. First,
an overview of what was done is presented under the title \textit{Data
  report overview}. Secondly, an index listing each variable along
with an indication of whether it was found to be problematic or not is
provided. Thirdly, as seen in Figure~\ref{fig:example2}, each variable
in the dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution when relevant and a list of flagged
issues, if any. For instance, as shown in Figure~\ref{fig:example2},
for the \code{numeric}-type variable \code{events} from
\code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and
\code{NaN}), while two values were also flagged as potential outliers
that should be investigated more carefully.

@@

\smallskip Control output and post-processing \\
\quad \code{output} & Type of output file to be produced (html, word (.docx) or pdf) & \code{NULL} (use pdf if \LaTeX\ is found, otherwise Word (if on Windows), or html)\\
\quad \code{render} & Should the output file be rendered from markdown? & \code{TRUE} \\
\quad \code{openResult} & If a  pdf/html file is rendered, should it
                       automatically open afterwards, and if not,
                       should the \code{rmarkdown} file be opened? & \code{TRUE} \\
\quad \code{replace} & Overwrite an existing file with the same name? & \code{FALSE} \\
\quad \code{vol} & Add a suffix to the file name of the outputted report &   \code{""} (no suffix)\\

@@

In summary, and as indicated in Figure~\ref{figure:cleanStructure}, there are two stages
where \code{makeDataReport()} applies functions to each of the variables:
\begin{enumerate}
\item In the pre-check stage.
\item As part of the summarize/visualize/check (SVC) steps.
\end{enumerate}
Each of these stages are controllable using appropriate function
arguments in \code{makeDataReport()}, and above we have shown examples
of how to tweak them to modify the data cleaning outputs. However, if
the dataset at hand requires new, additional checks, then more control
is needed. The package contains a vignette that explains the details of how to modify and
expand the possibilities by producing new summary, visual, and check
functions.

@@

Figure \ref{fig:bigExampleP45} presents the remaining two pages with variable presentations. On pages 4 and 5, we find a few remarks:
\begin{itemize}
\item In the \code{birthday} variable, there is an entry with the date March 1, 1300 which is a bit of an outlier. 
\item Among the states in which the presidents were born, New York was mistakenly spelled with a lower case "Y" in at least one entry.
\item The variable concerning assassination attempts is coded as a numeric, but the default \code{smartNum = TRUE} setting of \code{makeDataReport()} implies that such a numeric variable with only a few (less that 5) unique values is treated as a factor variable in the data report, thereby providing more relevant summaries, visualizations and checks. This is remarked in the variable presentation for \code{assassinationAttempt} and it can also be seen by the visualization being a barplot rather than a histogram.
\item The variable concerning the sex of the president was skipped, as there is nothing to present when all US presidents have been male so far.
\item The report flags the variable \code{ethnicity} to have suspiciously few observations in one category, "African American". 
\item A few presidents were found to have odd values in the variable describing the duration of their presidencies: Some had very short (outlier) presidencies of less than two years and one was registered to have an \textit{infinite} presidency. 
\item The variable concerning age at inauguration was coded as a character variable, but consists exclusively of numbers and takes a lot of different values and therefore, it was flagged as a potentially misclassified numeric variable.
\item It seems as if one or more presidents have complex numbers as their favorite numbers. As \code{complex} is not a supported variable type in \code{dataMaid} and no strategy for handling this class was provided in the \code{treatXasY} argument, the variable is simply flagged as non-supported.
\end{itemize}
A lot of these mistakes are easily fixable, and we will do so below. However, some of them require more delicate knowledge of the subject matter. For instance, \code{ethnicity} is very reasonably marked as a potentially problematic variable as it includes only a single observation of "African American". However, a human reading this report will know that this does \textit{not} reflect a mistake in the data, but rather a peculiarity in the real world, and as such, it should not be cleaned out.

@@

Next up is the unlikely US president birthday of March 1, 1300. In order to understand if this is a generally problematic observation, or if it is just a mistyped observation, we inspect the full data entry for this person. We can do this using the usual \proglang{R} selection syntax as above, or we can use the value of a \code{identifyOutliers} call to select this observation:

@@

We see that Obama is listed as being president forever, which history has proven to be wrong. Trump, on the other hand, has a missing value for his presidency duration, which is in fact reasonable as we cannot know how long it will be yet (anno 2017). Presidents William Harrison, Zachary Taylor and James Garfield were identified to have very brief presidencies, but these are not mistakes, as any US history textbook can tell us. Thus, the only mistake left to fix is Obama's infinite presidency:

@@

\subsection[Incorporating dataMaid in an automated workflow]{Incorporating \pkg{dataMaid} in automated workflow}
The default settings of \code{makeDataReport()} have been set to facilitate easy and quick data report generation, but unfortunately, this also means that it is not ideal for a more programming-oriented workflow, where the function might not be called by a human. For instance, one might be interested in automatically running \code{makeDataReport()} on all datasets received from a certain client, perhaps via email or through a web upload, and returning a data report for them to inspect and comment before ever looking at the data. In this scenario, there are a few issues with the standard data report:
\begin{enumerate}
\item After rendering, the report is automatically opened. This is not very useful, if the processes are supposed to run in the background.
\item The report generation writes messages to the console while producing and rendering the report.
\item Unless specifically told otherwise, every report created for the same dataset (or different datasets with the same storage name in \proglang{R}) will have the same file name.
\end{enumerate}
Please note that the data report does contain information about who, when and how concerning its generation, so even though the default choices for file names do not make it easy to tell different reports for the same dataset apart, it should be rather easy when inspecting the report manually.

The three problems can easily be solved by use of the arguments of \code{makeDataReport()}. Whether or not the outputted file is opened can be controlled through the argument \code{open}. How much information is printed in the console can be adjusted by using the argument \code{quiet}. And convenintely introducing small alterations of the file names can be obtained by use of the \code{vol} argument. For instance, we can make a data report for \code{toyData} that is not opened automatically, produces no output to the console and includes the date and time of its creation in the file name:

@@

\subsection[Using dataMaid for problem flagging]{Using \pkg{dataMaid} for problem flagging}
If the dataset is large and the time available for reading through the data report is scarce, it can be convenient to only make a report concerning the variables that were flagged to be problematic. This can be achieved by using the \code{onlyProblematic} argument for
\code{makeDataReport()}. By specifying \code{onlyProblematic = TRUE}, only
variables that raise a flag in the checking steps will be summarized
and visualized. But perhaps we are not even interested in obtaining
general information about these variables, but only in getting a quick
overview of the problems they might have. This is obtained by
using the \code{mode} argument:
